---
title: "Module 6 Lab"
output: pdf_document
---

The first part of this lab is in DataCamp. Complete Chapter 5, on **data frames**, in the "Introduction to R" course. Data frames are integral to data analysis in R; pay close attention.

The next two sections will cover **two sample t-tests** and **paired t-tests**. 

# Two Sample t-test

Suppose a marketing firm is testing the public approval of two new pairs of basketball shoes. They obtain two random samples of numeric consumer ratings of the shoes. We will randomly generate some make-believe samples, but let's set a seed first so we get the same results.

```{r}
set.seed(1964) # Nike was founded in 1964
shoeA <- rnorm(25, mean = 50, sd = 20)
shoeB <- rnorm(30, mean = 60, sd = 15)
```

Notice we have a different number of samples for each shoe; this is okay because it is **not** a paired t-test. We know the answer, but the shoe company wants to know whether or not the two shoes have the same average consumer rating. 
$$H_{0}: \mu_{A} - \mu_{B} = 0$$
$$H_{A}: \mu_{A} - \mu_{B} \neq 0$$
We will perform a t-test to find out.
```{r, results ='hide'}
t.test(shoeA, shoeB, mu = 0, alternative = "two.sided", paired = FALSE, var.equal = FALSE)
```

 The first two arguments are the sample observations. The final three arguments in `t.test()` here are the default, but I include them to highlight the details of our test. The third argument, `mu = 0`, declares that we want to test if the differnce of group means is equal to 0. Next, `alternative = "two.sided"` indicates that the alternative hypothesis is $H_{A}: \mu_{A} - \mu_{B} \neq 0$, rather than a one sided alernative. This is **not** a paired t-test, so we specify `paired = FALSE`. Finally, the shoe executives assume group variances are not the same; thus we use the argument `var.equal = FALSE`. In general, we do not know the individual group variances, and it is safer to assume they are not the same. In this context "safer" means that assuming group variances are the same when they are not could significantly affect results, whereas assuming they are different when they are in fact the same with have minimal impact.
 
```{r, echo = FALSE}
t.test(shoeA, shoeB, mu = 0, alternative = "two.sided", paired = FALSE, var.equal = FALSE)
```

The output is very similar to what we have seen before. The data is listed in the first line, followed by the test statistic, degrees of freedom, and p-value in the second line. The alternative hypothesis is stated next, followed by a 95\% confidence interval for $\mu_{A} - \mu_{B}$, and point estimates for the individual group means.

# Paired t-test

The shoe marketing executives also want to know if married husbands and wives feel the same way, on average, about shoeA. For this, they conduct a **paired t-test** on a sample of 32 randomly chosen couples.
$$H_{0}: \mu_{H-W} = 0$$
$$H_{A}: \mu_{H-W} \neq 0$$
Let's generate some make-believe data, pretending that there is no difference between paired husband and wife ratings, on average.
```{r}
set.seed(1809) # Carl Friedrich Gauss helps establish the Normal distribution in 1809
means <- rnorm(32, 50, 25) # Create 32 distinct means; one for each couple
wife <- rnorm(32, mean = means, sd = 15) 
husband <- rnorm(32, mean = means, sd = 15) # Give each married couple same mean
```
Notice that each married couple has their own distinct mean *shoeA* rating that they share. This means that, though the executives don't know it, the average difference in husband-wife rating is 0. Let's see what the test says. This time, we specifiy `paired = TRUE`, because we perform a paired t-test. Also, the executives think that married couples opinions vary by the same amount, so we will specifiy `var.equal = TRUE`.
```{r}
t.test(wife, husband, mu = 0, alternative = "two.sided", paired = TRUE)
```
By now the output should look familiar, so let's give a technical summary. 

Based on the results of a paired t-test on a random sample of 32 married couples, the executives fail to reject the null hypothesis that on average husbands and wives rate shoeA the same. The paired t-test had a test statistic of -0.8417, and a p-value of 0.4064. The executives estimate husbands rate shoeA 3.01 units lower than their wives, on average, with a 95\% confidence interval for this difference of -10.31 to 4.28.   



